---
title: "Analysis"
author: "Brent Cameron"
date: "12/15/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This script loads the processed data, does a simple analysis and saves the results to the results folder

# First, load needed packages
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidyverse) #for easier data manipulation
library(tidymodels) #for easier statistical analysis
```

# Path to Data
*note the use of the "here()" package and not absolute paths*
```{r}
data_location <- here::here("data","processed_data","processeddata.rds")

#load data. 
mydata <- readRDS(data_location)
```


# Data exploration/description
*Let's first examine a summary of our data before we continue with exploration*
```{r}
mysummary <- summary(mydata)
print(mysummary)
```

*As can be seen from the summary, all variables are considered characters except*
*for Deaths, which is considered **numerical**

**Please note...**
Our outcome of interest is Cancer deaths.
Variables that will serve as predictors of outcome include...
Age group, sex, total deaths, ethnicity, race, and cancer sites

Before we do plotting of the data comparing our predictors of interest, we 
will examine each variable separately to better understand the data we are 
working with

# First we will examine the age group of our data set:
```{r}
mydata %>% ggplot(aes(x = Age)) + geom_bar() +
  scale_x_discrete(guide = guide_axis(n.dodge = 1)) +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(angle = 90, hjust = 1)) 
```

As can be seen from the chart the population is largely composed of older 
adults between the ages of 50-85

# Now the sex of the data set:
```{r}
mydata %>% ggplot(aes(x = Sex)) + geom_bar()
```

The chart shows that this data set is almost exactly equal in terms of number
of male and female 

# Next we will examine the race of the data set:
```{r}
mydata %>% ggplot(aes(x = Race)) + geom_bar() +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

As can be seen from the data, we are working with a majority White population
with a moderate population of African American and a lower population of Asian or 
Pacific Islander and American Indian or Alaska Native

# Now ethnicity:
```{r}
mydata %>% ggplot(aes(x = Ethnicity)) + geom_bar() 
```

This chart shows us that the data is composed of a mostly non-Hispanic 
population, with Hispanic as the second most common followed by some unknown or
missing

# Finally cancer sites:
```{r}
mydata %>% ggplot(aes(x = CancerSite)) + geom_bar() +
  scale_x_discrete(guide = guide_axis(n.dodge = 1)) +
  theme(text = element_text(size = 8, face = "bold"),
        axis.text.x = element_text(angle = 90, hjust = 1))
```

It seems as though the largest number of cancer cases are localized to the 
digestive system, lung and bronchus, and respiratory system, with a large number 
of cases considered to be miscellaneous

Now that we have completed a preliminary look at our predictors of interest,
we will examine a closer look at the data to see if any basic trends 
can be seen, with the first plot, we will be examining cancer deaths and 
race to see if there exists any trends

# Plot 1, Plotting Cancer deaths by Race
```{r}
Cancer_Deaths_By_Race <- mydata  %>%
  ggplot(aes(x=Race, y=Deaths)) +
  geom_bar(stat = "identity") +
  ggtitle("Cancer Deaths by Race in the United States") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2, check.overlap = FALSE))+
  scale_y_continuous(labels = comma)

#Examine plot
print(Cancer_Deaths_By_Race)
```

As we can see from the chart, cancer mortality is largely concentrated to the 
White population group, with African American being the second highest

Now we will be plotting the number of Cancer deaths by Age Group to examine
If a trend is present between certain age groups having higher levels of 
relative mortality

# Plot 2, Plotting Cancer deaths by Age Group
```{r}
Cancer_Deaths_By_Age_Group <-mydata %>%
  ggplot(aes(x=Age, y=Deaths))+
  geom_bar(stat = "identity") +
  ggtitle("Cancer Deaths by Age Group in the United States")+
  geom_smooth(method='lm')+
  scale_y_continuous(labels = comma)+
  scale_x_discrete(guide = guide_axis(n.dodge = 1, check.overlap = TRUE))+
  theme(text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, hjust = 1)) 

#Examine the Plot
print(Cancer_Deaths_By_Age_Group)
```

Like the graph that was viewed earlier, deaths are largely concentrated towards
older populations

Now we will examine cancer deaths by site to see if there are particular 
sites that are responsible for a higher overall mortality

# Plot 3, Plotting Cancer Deaths by Site
```{r}
Cancer_Deaths_By_Site <- mydata %>%
  ggplot(aes(x= CancerSite, y=Deaths)) +
  geom_bar(stat = "Identity") +
  ggtitle("Cancer Deaths by Site")+
  geom_smooth(method='lm')+
  scale_x_discrete(guide = guide_axis(n.dodge = 1, check.overlap = TRUE))+
  scale_y_continuous(labels = comma)+
  theme(text = element_text(size = 6),
        axis.text.x = element_text(angle = 90, hjust = 1)) 

#Examine plot
print(Cancer_Deaths_By_Site)
```

Once again, as previously discussed in the prior chart examining count of 
cancer sites, deaths are concentrated in digestive, lung and bronchial,
and respiratory


#Plot 4, plotting cancer deaths by sex
```{r}
Cancer_Deaths_By_Sex <- Cancer_Data %>%
  ggplot(aes(x= Sex, y=Deaths)) +
  geom_bar(stat = "Identity") +
  ggtitle("Cancer Deaths by Sex")+
  geom_smooth(method='lm')+
  scale_x_discrete(guide = guide_axis(n.dodge = 1, check.overlap = TRUE))+
  scale_y_continuous(labels = comma)+
  theme(text = element_text(size = 6),
        axis.text.x = element_text(angle = 90, hjust = 1)) 

#Examine plot
print(Cancer_Deaths_By_Sex)
```






# summarize data 
```{r}
mysummary = summary(mydata)

#look at summary
print(mysummary)

#change shape to data frame (for easier saving/showing in manuscript)
summary_df = data.frame(do.call(cbind, lapply(mydata, summary)))

#save data frame table to file for later use in manuscript
summarytable_file = here("results", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Data fitting/statistical analysis
This section of the analysis script will focus on the statistical 
analysis of the data, through utilizing the "tidymodels" framework

Now that we have processed and explored the data, we will now 
work on splitting the data to create a training set and a testing set, 
which we will eventually use to measure model performance

First we fix the random numbers by setting the seed
This allows for the construction of a reproducible analysis when random numbers are utilized
```{r}
set.seed(123)

#We will now put 3/4 of our data into the training set
data_split <-initial_split(mydata, prop = 3/4)

#We will now create data frames for the two sets (training and testing)
train_data <- training(data_split)

test_data <- testing(data_split)
```

Now that we have data that is split into our training and test data, we will
now initiate a recipe to create a simple linear regression model
Note that our main continuous predictor will be cancer deaths, while our
main categorical predictors will be ethnicity and race 

we will start our models by creating a recipe for our first main categorical
predictor, which is ethnicity, and utilize all other predictors to determine
if any statistically significant associations exist
```{r}
mydata_rec <-
  recipe(Ethnicity ~ ., data = train_data) 

#Let's get the summary of our data to examine it more closely
summary(mydata_rec)
```


# Now that we have our recipe for our data, we will fit our model to the data
*Note that since we are using Ethnicity, which is categorical, we are using null_model() in place of linear regression*
```{r}
null_model <- null_model() %>%
  set_engine("parsnip") %>%
  set_mode("classification") %>%
  translate()
```


Now that we have set our model, we will use the "workflow" package to 
streamline the fitting process by bundling the recipe and models together
Note that we use our recipe in the process and not the data by itself
```{r}
mydata_wflow <-
  workflow() %>%
  add_model(null_model) %>%
  add_recipe(mydata_rec)
```

Now we will prepare the recipe and train the model
```{r}
mydata_fit <-
  mydata_wflow %>%
  fit(data = train_data)
```

Now we will examine our fitted model by using the 'extract_fit' function
```{r}
mydata_fit %>%
  extract_fit_parsnip() %>%
  tidy()
```

With the past commands, so far we have
**1. built our model**
**2. created a recipe**
**3. bundled the model with our recipe**
**4. trained our workflow using a single call to fit()**

Now we will use our trained workflow to predict with the test data using the 
predict() command, which applies the recipe to the new data then passes the
collected information to the fitted model
```{r}
predict(mydata_fit, test_data)
```


We will use the 'augment' function to predict each possibility of our data
```{r}
mydata_aug <- augment(mydata_fit, test_data)

mydata_aug %>%
  select(Ethnicity, .pred_class)
```


Now we will examine the relationship between Ethnicity as a predictor for the training data
```{r}
as.factor(mydata_aug$Ethnicity)
predict(mydata_fit, train_data)
mydata_aug <- augment(mydata_fit, train_data)
mydata_aug %>%
  select(Ethnicity, .pred_class)
```


Finally, we will examine an alternative model with only the main predictor 
Deaths, to Ethnicity, rather than the other chosen predictors
We must create a new recipe before we continue
```{r}
alternative_rec <-
  recipe(Ethnicity ~ Deaths, data = train_data) 

#Just like before, we will create a workflow with the recipe
alternative_rec_workflow <-
  workflow() %>%
  add_model(null_model) %>%
  add_recipe(alternative_rec)

#Now we will prepare the recipe and train the model
alternative_fit <-
  alternative_rec_workflow %>%
  fit(data = train_data)

#Now we will examine our fitted model by using the 'extract_fit' function
alternative_fit %>%
  extract_fit_parsnip() %>%
  tidy()

```

# save fit results table  
```{r}
Ethnicity_Deaths_Table=here("results", "resulttable.rds")
saveRDS(Ethnicity_Deaths_Table, file = "Ethnicity_Deaths_Table")
```